---
title: "Janitor exercises"
author: "Abigail Peña Alejos & Nikolina Klatt"
date: "2022-11-13"
output: 
 html_document:
    toc: TRUE
    df_print: paged
    number_sections: FALSE
    highlight: tango
    theme: lumen
    toc_depth: 3
    toc_float: true
    css: custom.css 
    self_contained: false
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today we will check some of the basic functions offered by the `janitor package` that will help us save time when cleaning our data. Along with `janitor`, we will make use of `tidyverse`packages such as `dplyr` and `readr`. `kableExtra` is suggested to render nice tables. But it is up to you! 

```{r, message = F}
library(readr)
library(tidyverse)
library(janitor)
library(kableExtra)
```

For this exercise we will use *crime_subset.csv* dataset taken from a real world data hosted at [Data.gov catalogue](https://catalog.data.gov/dataset), the U.S. Government’s open data portal. The dataset contains reported crimes in the Montgomery County in Maryland. (You can download the original [here](https://catalog.data.gov/dataset/crime)) 

We import the .csv file 

```{r, warning=FALSE, message=FALSE}

crime_df <- read_delim("data/crime_subset.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

```

Now, let's go!

***


## Getting our variables names in clean format with `clean_names()`

We have talked in our presentation about naming convention in computing, for example for variable and subroutine names, and for file names. We have been recommended using `snake_case`, the style of writing in which each whitespace is replaced by an underscore `_` character, and each word is written in lowercase. Something like this: `my_awesome_df`.

However, this isn't the case of our dataset. 

```{r}
crime_df %>% 
  glimpse()

```

**EXERCISE 1:** Change variable names of the dataframe to follow `snake_case` convention.

```{r}
crime_df <- crime_df %>% 
              clean_names()
crime_df %>% 
  names()  

```
Voilá! Now we have `snake_case` formatted variables! Btw, did you notice the function also took care of the `/` in `Dispatch/Date` and replaced it with an underscore. Isn't it great?

Let's pretend you are not really fond of `snake_case` format because you prefer the `camelCase`convention instead. You might also want to leave NIBRS and ID untouched. No problem! `clean_names()` got your back. Just change your preference in the argument `case =` to `upper_camel` and keep the abbreviation with the argument `abbreviations`. Check out `?clean_names()` for further information.

```{r}
crime_df %>% 
  clean_names(case = "upper_camel", abbreviations = c("ID", "NIBRS")) %>% 
  names()
```

## Detect duplicates with `get_dupes()`

**EXERCISE 2:** Find out whether there are any duplicate values in the *crime_df* dataframe

```{r}
crime_df %>% 
  get_dupes(crime_name2,crime_name1,victims, cr_number) %>% 
  arrange(desc(dupe_count))

```
Uhlala, seems like our data needs some cleaning!


## Remove content with `remove_empty()` and `remove_constant()` functions


Our `crime_df` data contain several missing values, and a couple of rows/columns filled with NAs.

**Exercise 3.1:** Create a subset called *tidy_crime* using the `remove_empty()`

```{r}
tidy_crime <-  crime_df %>% 
               remove_empty(c("rows","cols"), quiet = F)

glimpse(tidy_crime)
```

**EXERCISE 3.2:** From `crime_df`, subset the observation for crimes committed in *Silver Spring district*, clean the data and get rid of the *police_distric_name* variable. 

```{r}

s_silver_spring <-  crime_df %>% 
  filter(police_district_name == "SILVER SPRING") %>% 
  remove_empty(c("rows","cols"), quiet = F) %>% 
  remove_constant(quiet = F)

crime_df %>% 
  filter(police_district_name == "SILVER SPRING") %>% 
  remove_empty(c("rows","cols"), quiet = F) %>% 
  remove_constant(quiet = F) %>% 
  glimpse()

```

## Check content of columns with `compare_df_colums`

`compare_df_colums` is a pretty cool function to check how a group of dataset compare to each other and whether could be possible to merge them based on the number of variables and vector class. 

**Exercise 4** From *crime_df*, subset the observation for crimes committed in *Wheaton district* , clean the data and get rid of the *police_distric_name* variable. Then compare *Wheaton* and *Silver Spring* subsets. Check out what columns are missing or present in the different inputs. 


```{r}

s_wheaton <- crime_df %>% 
  filter(police_district_name == "WHEATON") %>% 
  remove_empty(c("rows","cols"), quiet = F) %>% 
  remove_constant(quiet = F)

compare_df_cols(s_silver_spring, s_wheaton)
```

## Using `tabyl()` and `adorn_*()` options

We know that `tabyl()`is useful to create contingency tables with  1, 2 or 3 variables, and we have learned that `adorn_*()` functions will help us add information such as total numbers, not to forget a  nice formatting for percentages.

**EXERCISE 5.1:** Find out proportion of crime per district, round percentages to 1 digit.  

```{r}
  
tabyl(tidy_crime$police_district_name) %>% 
  adorn_pct_formatting(digits = 1) 

```

**EXERCISE 5.2:** For crimes committed in Silver Springs involving only one victim, find what percentage corresponds to property-related offences. DON'T USE `drop_na()`
```{r}

s_silver_spring %>% 
tabyl(crime_name1, victims) %>%
adorn_totals("row") %>%
adorn_percentages("col")  %>%
adorn_pct_formatting(digits= 1)  %>%
adorn_ns()  %>%
adorn_title() 

```
## Finally, let's play with numbers & dates

**Rounding up numbers with `round_half_up`**

Following the example of *v1* vector, create 2 vectors in which the numbers from vector 1 are rounded up to 1 and 0 digits respectively. Bind the vectors into a tibble under the name `funny_numbers` to see how the numbers compare across columns

```{r}
v1 <- runif(10, min=15, max=48)
v2 <- round_half_up(v1, digits = 1)
v3 <- round_half_up(v1)
  
funny_numbers <- tibble(v1,v2,v3)
funny_numbers
```


**Playing with dates**

Use the `excel_numeric_to_date()` function to convert Excel serial date numbers in *edates* to class `Date`

```{r}

edates <- c(0, 59, 61, 1000, 45100)

edates %>% 
  excel_numeric_to_date()

```

## Sources

[janitor package](https://cran.r-project.org/web/packages/janitor/janitor.pdf)

[overview of janitor functions](https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html#tabyl---a-better-version-of-table) 

[cleaning penguins with the janitor packages](http://jenrichmond.rbind.io/post/digging-into-the-janitor-package/)

[How to Clean Data: {janitor} Package](https://www.exploringdata.org/post/how-to-clean-data-janitor-package/)
